Reflections on reflection for \\ intuitionistic set theories

### 2019-01-11 paper-reflection.tex
to do
Fix critical error

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\DeclareMathOperator{\rank}{rank}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

\newcommand{\PRA}{\textsc{pra}}
\newcommand{\ZF}{\textsc{zf}}
\newcommand{\IZF}{\textsc{izf}}
\newcommand{\CZF}{\textsc{czf}}
\newcommand{\ZFC}{\textsc{zfc}}
\newcommand{\ZFCS}{\textsc{zfc/s}}
\newcommand{\ZFCI}{\textsc{zfc+i}}
\newcommand{\RRS}{\textsc{rrs}}
\newcommand{\RDC}{\textsc{rdc}}
\newcommand{\DC}{\textsc{dc}}
\newcommand{\MDC}{\textsc{mdc}}
\newcommand{\ES}{(\IZF+\RRS)/\textsc{s}}

@@

\begin{abstract}
  We study the well-known reflection principle of Zermelo--Fraenkel set theory
  in the context of intuitionistic Zermelo--Fraenkel set theory~$\IZF$. We show
  that the reflection principle is equivalent to Aczel's relation reflection scheme~$\RRS$. As
  applications, we give a new proof that relativized dependent choice is
  equivalent to the conjunction of the relation reflection scheme and dependent
  choice and we present an intuitionistic version of Feferman's~$\ZFCS$, a
  conservative extension of~$\ZFC$ which is useful as a foundation for category
  theory.
\end{abstract}

@@

\begin{thm}The reflection principle is equivalent, over~$\IZF$, to Aczel's relation reflection
scheme~$\RRS$.\end{thm}

Aczel's relation reflection scheme will be reviewed below. It is validated not
only by~$\ZFC$, but also by~$\ZF$, and furthermore by
all known models of~$\IZF$, hence might be regarded as not entirely
unconstructive, even though it's conjectured to be independent of~$\IZF$. As a
result, the question whether the reflection principle holds for~$\IZF$ remains
open (though conjectured to be false), and for the stronger system~$\IZF+\RRS$
we can give a variant ``with smallness'' which can serve as a set-theoretic
foundation for category theory.

@@

\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS$ by
adding a constant symbol~$\SS$ together with axioms
stating that~$\SS$ is transitive, closed under subsets and reflective for all
formulas of the original language (that is, adding one instance
of~\eqref{ax:refl} for each formula of~$\IZF$).\end{defn}

\begin{cor}The system~$\ES$ is conservative
over~$\IZF+\RRS$.\end{cor}

This note is organized as follows. Section~\ref{sect:review} reviews the
classical proof of the reflection principle in the context of~$\ZF$ set theory.
Section~\ref{sect:classical-proof} reviews Aczel's relation reflection scheme
and presents our main result. We conclude in Section~\ref{sect:appl-rdc} and
Section~\ref{sect:appl-smallness} with two short applications.

@@

A basic proof of the reflection principle for~$\ZF$ runs as follows. Our proof
of the reflection principle for~$\IZF+\RRS$ will follow the same outline.

@@

\begin{proof}Let a set~$M$ be given. We obtain~$S$ by applying
Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$. That
the resulting set~$S$ has the required property can be checked by an induction on the structure
of~$\varphi$. The cases~$\text{``$=$''}, \text{``$\in$''}, \text{``$\top$''},
\text{``$\bot$''}, \text{``$\wedge$''},
\text{``$\vee$''}, \text{``$\Rightarrow$''}$ follow
trivially from the induction hypothesis. The case~``$\forall$'' does
not need to be treated since we may assume without loss of generality that all
universal quantifiers in~$\varphi$ have been rewritten
as~``$\neg\exists\neg$''.

@@

@@

Our constructive rendition of the reflection principle will require Aczel's
relation reflection scheme~$\RRS$ displayed in Definition~\ref{defn:rrs}. This
result is the best possible, as we verify in Theorem~\ref{thm:refl-entails-rrs}
that conversely the reflection principle entails the
relation reflection scheme. The relation reflection scheme first surfaced in
the theory of coinductive definitions of classes and enjoys substantial
stability properties, as it passes from the meta theory to XXX[all kinds of]
models. Background on~$\RRS$ can be found in Aczel's original article
introducing it~\cite{aczel:rrs}.

\begin{defn}[$\RRS$, Aczel's relation reflection scheme]\label{defn:rrs}
Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a set~$B$ such that~$A \subseteq B \subseteq X$ and such
that~$\forall x \in B\_ \exists y \in B\_ \langle x,y \rangle \in R$.
\end{defn}

For our purposes, Palmgren's multivalued dependent choice is slightly more
convenient to work with than~$\RRS$. It is equivalent, over~$\CZF$ and
a~fortiori over~$\IZF$, to~$\RRS$~\cite{palmgren:mdc}.

\begin{defn}[$\MDC$, Palmgren's multivalued dependent choice]
Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$A \subseteq X$ be a
subset. If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a function~$f : \NN \to P(X)$ (the class of all subsets of~$X$)
such that~$A \subseteq f(0)$ and such that~$\forall x \in f(k)\_ \exists y \in
f(k+1)\_ \langle x,y \rangle \in R$ for every number~$k$.\end{defn}

@@

\begin{lemma}\label{lemma:izf-smallstep}
Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\begin{multline*}
  {\qquad}\forall M\_
  \exists S \supseteq M\_
  \forall u_1,\ldots,u_n \in S\_ \\
  \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in S\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in S\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
  {\qquad}
\end{multline*}
Furthermore: (1)~We may suppose that~$S$ is transitive. (2)~We may suppose
that~$S$ is closed under subsets. (3)~Given a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of the single
formula~$\varphi$, we may suppose that~$S$ bounds all of the
formulas~$\varphi_i$.
% XXX: better explanation
\end{lemma}

\begin{proof}We first consider the case~$n = 1$.

By~$\MDC$ and Lemma~\ref{lemma:izf-microstep}, there is a
function~$f : \NN \to P(V)$, where~$V$ is the universe, such that~$M \in
f(0)$ and such that for any number~$k$ and any set~$H \in f(k)$, there is a
set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
Lemma~\ref{lemma:izf-microstep}.

We set~$S \defeq \bigcup \bigl(\bigcup_{k \in \NN} f(k)\bigr)$. This set~$S$
has the required property.

In order to ensure addendum~(1) and~(2), we apply~$\MDC$ in a slightly
different way such that for any number~$k$ and any set~$H \in f(k)$, there is a
set~$H' \in f(k+1)$ such that~$H$ and~$H'$ are related as in the conclusion of
Lemma~\ref{lemma:izf-microstep} and such that~$H'$ is transitive and closed
under subsets. Even though it cannot be expected that for any number~$k$,
any set~$H \in f(k)$ will be transitive and closed under subsets, the union~$S$
will. Addendum~(3) can be ensured because of addendum~(3) of
Lemma~\ref{lemma:izf-microstep}.

For the case~$n \neq 1$, \textbf{???}
\end{proof}

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
under subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

\begin{proof}The proof of Theorem~\ref{thm:zf-bigstep} carries over. The only
difference is that instead of Lemma~\ref{lemma:zf-smallstep},
Lemma~\ref{lemma:izf-smallstep} has to be used, and that the case for the
universal quantifier has to be treated just as the case for the existential
quantifier is.\end{proof}

\begin{thm}\label{thm:refl-entails-rrs}
Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS$ can be
deduced from suitable instances of the assumption that, given a finite list
of formulas, for every set~$M$ there is a set~$S \supseteq M$ reflecting the
given formulas.
\end{thm}

\begin{proof}We verify the equivalent~$\MDC$ instead of~$\RRS$. Let~$X$ and~$R
\subseteq X \times X$ be classes. Let~$A \subseteq X$ be a subset and
suppose~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$.

By assumption, there is a set~$S \supseteq A$ which reflects the three
formulas~``$x \in X$'', ``$\langle x,y \rangle \in R$'' and~``$\exists y \in X\_ \langle
x,y \rangle \in R$''.

By separation, the class~$S \cap X$ is a set, and hence we may set~$f(k) \defeq S \cap X$
for all numbers~$k$. Then~$A \subseteq f(0)$; any given~$x \in f(k)$, by
assumption and the reflecting property of~$S$ there exists~$y \in S$ such
that~$y \in X$ (hence~$y \in f(k+1)$) and~$\langle x,y \rangle \in R$.
\end{proof}

@@

\begin{defn}[$\DC$, dependent choice]
Let~$X$ and~$R \subseteq X \times X$ be sets. Let~$x_0 \in X$.
If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
\end{defn}

\begin{defn}[$\RDC$, relative dependent choice]
Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$.
If~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$,
then there is a function~$f : \NN \to X$ such that~$f(0) = x_0$ and
such that~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
\end{defn}

@@

\begin{proof}It's clear that~$\RDC$ implies~$\RRS + \DC$. % XXX: check

@@

By Theorem~\ref{thm:izf-bigstep}, reflection is available; hence there is a
set~$S \supseteq \{x_0\}$ which reflects the three formulas~``$x \in X$'', ``$\langle x,y
\rangle \in R$'' and~``$\forall x\_ x \in X \Rightarrow \exists y\_ (y \in X \wedge \langle x,y
\rangle \in R)$''.

@@

\section{An intuitionistic version of Feferman's~$\ZFCS$}

@@

However, in the context of~($\IZF$ or)~$\ZFC$, it is difficult to make this
claim precise. Because in~$\ZFC$ the category~$\Set$ of all sets can not be
coded as a set,~$\ZFC$ cannot define truth in~$\Set$. We must therefore resort
to a meta theory in order to express this claim, for instance by stating
that primitive recursive arithmetic~$\PRA$ proves that for any
formula~$\varphi$ of~$\ZFC$, $\ZFC$ proves that~$(\Set \models \varphi)
\Leftrightarrow \varphi$, where~``$\Set \models \varphi$'' is to be unrolled
by~$\PRA$.

An alternative is offered by~$\ZFCI$, $\ZFC$ plus the existence of a
strongly inaccessible cardinal. In this system, there is a Grothendieck
universe~$U$, we can form the category~$\Set_U$ of all sets of~$U$ as an honest
set, can define truth in~$\Set_U$ and prove, within the system, that for any
formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
\varphi)$. This even holds for formulas of the full infinitary language of
toposes, which allows for infinite disjunctions and infinite conjunctions; this
could not be mimicked by resorting to~$\PRA$ as indicated above.

@@

% XXX: Reference more choice axioms?

>>>>

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\SS}{\mathbb{S}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\DeclareMathOperator{\rank}{rank}
\DeclareMathOperator{\length}{length}
\DeclareMathOperator{\List}{List}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

\newcommand{\PRA}{\textsc{pra}}
\newcommand{\ZF}{\textsc{zf}}
\newcommand{\IZF}{\textsc{izf}}
\newcommand{\CZF}{\textsc{czf}}
\newcommand{\ZFC}{\textsc{zfc}}
\newcommand{\ZFCS}{\textsc{zfc/s}}
\newcommand{\ZFCI}{\textsc{zfc+i}}
\newcommand{\RRS}{\textsc{rrs}}
\newcommand{\RDC}{\textsc{rdc}}
\newcommand{\DC}{\textsc{dc}}
\newcommand{\MDC}{\textsc{mdc}}
\newcommand{\ES}{(\IZF+\RRS_2)/\textsc{s}}

@@

\begin{abstract}
  We study the well-known reflection principle of Zermelo--Fraenkel set theory
  in the context of intuitionistic Zermelo--Fraenkel set theory~$\IZF$. We show
  that the reflection principle is equivalent to~$\RRS_2$, a strengthened
  version of Aczel's relation reflection scheme. As
  applications, we give a new proof that relativized dependent choice is
  equivalent to the conjunction of the relation reflection scheme and dependent
  choice and we present an intuitionistic version of Feferman's~$\ZFCS$, a
  conservative extension of~$\ZFC$ which is useful as a foundation for category
  theory.
\end{abstract}

@@

\begin{thm}The reflection principle is equivalent, over~$\IZF$, to the strong
relation reflection scheme~$\RRS_2$.\end{thm}

We also give a weaker version of the reflection principle which is equivalent
to Aczel's original reflection scheme~$\RRS$.
Both~$\RRS$ and~$\RRS_2$ will be reviewed below. They are validated not
only by~$\ZFC$, but also by~$\ZF$, and furthermore by
all known models of~$\IZF$, hence might be regarded as not entirely
% XXX: be more precise in "all known"
unconstructive, even though they are conjectured to be independent of~$\IZF$. As a
result, the question whether the reflection principle holds for~$\IZF$ remains
open (though conjectured to be false), and for the stronger system~$\IZF+\RRS_2$
we can give a variant ``with smallness'' which can serve as a set-theoretic
foundation for category theory.

@@

\begin{defn}The system $\ES$ is obtained from~$\IZF+\RRS_2$ by
adding a constant symbol~$\SS$ together with axioms
stating that~$\SS$ is transitive, closed under subsets and reflective for all
formulas of the original language (that is, adding one instance
of~\eqref{ax:refl} for each formula of~$\IZF$).\end{defn}

\begin{cor}The system~$\ES$ is conservative
over~$\IZF+\RRS_2$.\end{cor}

This note is organized as follows. Section~\ref{sect:review} reviews the
classical proof of the reflection principle in the context of~$\ZF$ set theory.
Section~\ref{sect:axioms} reviews Aczel's relation reflection scheme and its
variants. Our main result is presented in Section~\ref{sect:constructive-proof}.
We conclude in Section~\ref{sect:appl-rdc} and
Section~\ref{sect:appl-smallness} with two short applications.

@@

A basic proof of the reflection principle in~$\ZF$ runs as follows. Our proof
of the reflection principle in~$\IZF+\RRS_2$ will follow the same outline.

@@

\begin{proof}Let a set~$M$ be given. We obtain~$S$ by applying
Lemma~\ref{lemma:zf-smallstep} to the list of all subformulas of~$\varphi$. That
% XXX: "all" subformulas
the resulting set~$S$ has the required property can be checked by an induction on the structure
of~$\varphi$. The cases~$\text{``$=$''}, \text{``$\in$''}, \text{``$\top$''},
\text{``$\bot$''}, \text{``$\wedge$''},
\text{``$\vee$''}, \text{``$\Rightarrow$''}$ follow
trivially from the induction hypothesis. The case~``$\forall$'' does
not need to be treated since we may assume without loss of generality that all
universal quantifiers in~$\varphi$ have been rewritten
as~``$\neg\exists\neg$''.

@@

\section{Aczel's relation reflection axiom scheme and its variants}
\label{sect:axioms}

We write~``$R : X \rightrightarrows Y$'' to mean~``$\forall x \in X\_ \exists y
\in Y\_ \langle x,y \rangle \in R$''.

\begin{defn}Let~$X$ and~$R \subseteq X \times X$ be classes. Let~$x_0 \in X$.
\begin{itemize}
\setlength{\itemsep}{0.3em}
\item[$\DC$] \emph{dependent choice} \\ If~$X$ and~$R$ are sets and
if~$R : X \rightrightarrows X$, then there is a function~$f : \NN \to X$ such that~$f(0) =
x_0$ and~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
\item[$\RDC$] \emph{relativized dependent choice} \\ If~$R : X
\rightrightarrows X$, then there is a function~$f : \NN \to X$ such that~$f(0)
= x_0$ and~$\langle f(k), f(k+1) \rangle \in R$ for all numbers~$k$.
\item[$\RRS$] \emph{Aczel's relation reflection scheme} \\
If~$R : X \rightrightarrows X$, then there is
a set~$B$ such that~$x_0 \in B \subseteq X$ and~$R : B \rightrightarrows B$.
\item[$\MDC$] \emph{Palmgren's multivalued dependent choice} \\
If~$R : X \rightrightarrows X$, then there is a function~$f : \NN \to P(X)$
(the class of all subsets of~$X$) such that~$x_0 \in f(0)$ and such
that~$\forall x \in f(k)\_ \exists y \in f(k+1)\_ \langle x,y \rangle \in R$
for every number~$k$.
\end{itemize}
\end{defn}

\begin{defn}\label{defn:rrs2}
Let~$X$ and~$R \subseteq X \times X \times X$ be classes. Let~$x_0 \in X$.
\begin{itemize}
\setlength{\itemsep}{0.3em}
\item[$\RRS_2$]
If~$R : X \times X \rightrightarrows X$, then there is a set~$B$ such that~$x_0
\in B \subseteq X$ and~$R : B \times B \rightrightarrows B$.
\item[$\MDC_2$]
If~$R : X \times X \rightrightarrows X$, then there is a function~$f : \NN \to
P(X)$ such that~$x_0 \in f(0)$ and such that~$\forall x \in f(k)\_ \forall x'
\in f(k')\_ \exists y \in f(\max\{k,k'\}+1)\_ \langle x,x',y \rangle \in R$ for
any numbers~$k, k'$.
\end{itemize}
\end{defn}

Aczel's relation reflection scheme~$\RRS$ first surfaced in
the theory of coinductive definitions of classes and enjoys substantial
stability properties, as it passes from the meta theory to XXX[all kinds of]
models. Background on~$\RRS$ can be found in Aczel's original article
introducing it~\cite{aczel:rrs}. $\RRS$ is equivalent, over~$\CZF$ and
a~fortiori over~$\IZF$, to Palmgren's multivalued dependent
choice~\cite{palmgren:mdc}; this observation makes the relationship to
dependent choice more visible.

\begin{rem}Most published renderings of~$\RRS$ and~$\MDC$ do not refer to an initial
element~$x_0 \in X$, but to a set~$A \subseteq X$ and then require instead
of that~$x_0 \in B$, that~$A \subseteq B$. This difference is immaterial, thanks in
one direction to the existence of singleton sets and in the other to strong
collection. The same is true for~$\RRS_2$ and~$\MDC_2$, though the proof is not
as easy.
\end{rem}

% First prove that in RRS_2^∈, we may suppose that not only a single given
% element x0 is contained in B, but also a further given element x1.
% Do this by applying RRS_2^∈ to X^ := { ∅ } U { {x} | x ∈ X } and
% R^(p,p',q) given by:
% * p = ∅   and p' = ∅    and q = {x0}  or
% * p = ∅   and p' = {x0} and q = {x1}  or  the other way round  or
% * p = {x} and p' = {x'} and q = {y} and R(x,x',y).
%
% Now let R : X × X ⇉ X and a set A ⊆ X be given. We want to verify RRS_2^⊆
% for this data.
%
% Given H ∈ P(X), H' ∈ P(X), consider:
%
%    Let x ∈ H, x' ∈ H'.
%    Then by the strengthened RRS_2^∈, there is a set B ⊆ X
%    such that x, x' ∈ B and R : B × B ⇉ B.
%
% We can use collection to obtain a set J of such B's, and consider U J.
%
% We can then apply RRS_2^∈ to this relation to obtain a set T such that
% A ∈ T ⊆ P(X) and ... Set C := U T. Then A ⊆ C ⊆ X and R : C × C ⇉ C.

Aczel's $\RRS$ should not in itself be regarded as a choice principle; however in
its presence, ordinary~$\DC$ entails~$\RDC$. This result is due to
Aczel~\cite[Theorem~2.4]{aczel:rrs}, who proved the equivalence~$\RDC = \RRS +
\DC$ over~$\CZF^-$, and we give a new proof of this equivalence over the
stronger theory~$\IZF$, in Section~\ref{sect:appl-rdc}.

In the presence of~$\DC$, $\RRS$ is equivalent to~$\RRS_2$:

\begin{prop}\label{prop:rdc-rrs2-dc}
Over~$\CZF^-$, $\RDC$ is equivalent to~$\RRS_2 + \DC$.
\end{prop}

\begin{proof}Trivially,~$\RDC$ entails~$\DC$, and~$\RRS_2 + \DC$
entails~$\RDC$ by Aczel's result since~$\RRS_2$ entails~$\RRS$.

To verify that~$\RDC$ entails~$\RRS_2$, let classes~$X$ and~$R \subseteq X
\times X \times X$ be given, let~$x_0 \in X$ be an element and assume~$R : X
\times X \rightrightarrows X$. Let~$\List(X)$ be the class of finite lists with
entries in~$X$. We declare a class~$\widehat X$ by
\[ \widehat X \defeq \{ \langle i,j,v \rangle \,|\, i,j \in
\NN, v \in \List(X), i,j < \length(v) \} \]
and a relation~$\widehat R \subseteq \widehat X \times \widehat X$ by
defining~$\langle \langle i,j,v \rangle, \langle i',j',v' \rangle \rangle \in
\widehat R$ to be equivalent to
\begin{indentblock}
there exists an element~$y \in X$ such that
\begin{enumerate}
\item $\langle v!i, v!j, y \rangle \in R$ (where~$v!k$ is the element of the
list~$v$ at position~$k$),
\item $v'$ is obtained from~$v$ by adding the single element~$y$ at the end and
\item $\langle i',j' \rangle$ is the next point after~$\langle i,j
\rangle$ in some fixed enumeration of~$\NN^2$ which, for each number~$n$, first visits
all points~$\langle k,l \rangle$ with~$k,l < n$ before it visits any of
the other points.
\end{enumerate}
\end{indentblock}
Then~$\widehat R : \widehat X \rightrightarrows \widehat X$. Applying~$\RDC$
with initial value~$\langle 0,0, [x_0] \rangle \in \widehat X$ yields a
function~$f : \NN \to \widehat X$. Let~$B$ be the set of all
entries of the lists contained in the tuples~$f(k)$. Then~$x_0 \in B
\subseteq X$ and~$R : B \times B \rightrightarrows B$.
\end{proof}

A consequence of Proposition~\ref{prop:rdc-rrs2-dc} is that~$\RRS_2$ holds in
Aczel's type-theoretic ``sets as trees'' model
of~$\CZF$~\cite{aczel:sets-as-trees}, since that model validates~$\RDC$.

\begin{prop}Over~$\CZF^-$, $\RRS_2$ is equivalent to~$\MDC_2$.\end{prop}

\begin{proof}The proof in~\cite{palmgren:mdc} carries over.\end{proof}

% XXX: RRS_2 holds in ZF.

@@

Our constructive rendition of the reflection principle will require the axiom
scheme~$\RRS_2$ displayed in Definition~\ref{defn:rrs2}. This result is the
best possible, as we verify in Theorem~\ref{thm:refl-entails-rrs2} that
conversely the reflection principle entails~$\RRS_2$.

@@

\begin{lemma}\label{lemma:izf-smallstep}
Let~$\varphi(u_1,\ldots,u_n,x)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
\begin{multline*}
  {\qquad}\forall M\_
  \exists S \supseteq M\_
  \forall u_1,\ldots,u_n \in S\_ \\
  \begin{array}{@{}rcl@{}}
  (\exists x\_ \varphi(u_1,\ldots,u_n,x)) &\Longrightarrow&
  (\exists x \in S\_ \varphi(u_1,\ldots,u_n,x)) \quad\mathop{\wedge} \\
  (\forall x\_ \varphi(u_1,\ldots,u_n,x)) &\Longleftarrow&
  (\forall x \in S\_ \varphi(u_1,\ldots,u_n,x)).\end{array}
  {\qquad}
\end{multline*}
Furthermore: (1)~We may suppose that~$S$ is transitive. (2)~We may suppose
that~$S$ is closed under subsets. (3)~Given a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas instead of the single
formula~$\varphi$, we may suppose that~$S$ bounds all of the
formulas~$\varphi_i$.
% XXX: better explanation
\end{lemma}

\begin{proof}By~$\RRS_2$, there is a set~$B$ such that~$M \in B$ and such that
for any~$H_1, H_2 \in B$, there is a set~$H' \in B$ such that~$H \defeq H_1
\cup H_2$ and~$H'$ are related as in the conclusion of Lemma~\ref{lemma:izf-microstep}.

We set~$S \defeq \bigcup B$. This set has the required property. To verify
this, it is useful to observe that given~$u_1,\ldots,u_n \in S$, there is a
common set~$H \in B$ such that~$u_1,\ldots,u_n \in H$.

In order to ensure addendum~(1) and~(2), we apply~$\RRS_2$ in a slightly
different way to guarantee that for any sets~$H_1, H_2 \in B$, there is a
set~$H' \in B$ such that~$H \defeq H_1 \cup H_2$ and~$H'$ are related as in the
conclusion of Lemma~\ref{lemma:izf-microstep} and such that furthermore~$H'$ is
transitive and closed under subsets. Even though it cannot be expected that any
particular set~$H \in B$ will be transitive and closed under subsets, the
union~$S$ will. Addendum~(3) can be ensured because of addendum~(3) of
Lemma~\ref{lemma:izf-microstep}.
\end{proof}

\begin{rem}In the special case~$n = 1$, the proof of Lemma~\ref{lemma:izf-smallstep} can
be simplified to only use~$\RRS$ instead of~$\RRS_2$.\end{rem}

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS_2$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
under subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

\begin{proof}The proof of Theorem~\ref{thm:zf-bigstep} carries over. The only
difference is that instead of Lemma~\ref{lemma:zf-smallstep},
Lemma~\ref{lemma:izf-smallstep} has to be used, and that the case for the
universal quantifier has to be treated just as the case for the existential
quantifier has to.\end{proof}

% XXX: State mini-reflection theorem.

\begin{thm}\label{thm:refl-entails-rrs2}
Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS_2$ can be
deduced from suitable instances of the assumption that, given a finite list
of formulas, for every set~$M$ there is a set~$S \supseteq M$ reflecting the
given formulas.
\end{thm}

\begin{proof}Let~$X$ and~$R \subseteq X \times X \times X$ be classes. Let~$x_0 \in X$
be an element and suppose~$R : X \times X \rightrightarrows X$.

By assumption, there is a set~$S \ni x_0$ which reflects the three
formulas~``$x \in X$'', ``$\langle x,x',y \rangle \in R$'' and~``$\exists y \in X\_ \langle
x,x',y \rangle \in R$''.

The class~$B \defeq X \cap S$ is a set by separation and contains~$x_0$.
Given~$x,x' \in B$, there is a set~$y$ such that~$y \in X$ and~$\langle x,y,y'
\rangle \in R$. By the reflecting property of~$S$, we can assume that such an
element~$y$ exists in~$S$.

Hence~$R : B \times B \rightrightarrows B$.
\end{proof}

@@

@@

\begin{proof}Trivially,~$\RDC$ implies~$\DC$, and~$\RDC$ implies~$\RRS$ by a
similar, though much simpler, argument as in the proof of
Proposition~\ref{prop:rdc-rrs2-dc}.

@@

\textbf{XXX: Use mini-reflection theorem instead.}
By Theorem~\ref{thm:izf-bigstep}, reflection is available; hence there is a
set~$S \supseteq \{x_0\}$ which reflects the three formulas~``$x \in X$'', ``$\langle x,y
\rangle \in R$'' and~``$\forall x\_ x \in X \Rightarrow \exists y\_ (y \in X \wedge \langle x,y
\rangle \in R)$''.

@@

\section{An intuitionistic version of Feferman's~$\ZFCS$}
\label{sect:appl-smallness}

@@

However, in the context of~($\IZF$ or)~$\ZFC$, it is difficult to make this
claim precise. Because in~$\ZFC$ the category~$\Set$ of all sets can not be
coded as a set,~$\ZFC$ cannot define truth in~$\Set$. We must therefore resort
to a meta theory in order to express this claim, for instance by stating
that primitive recursive arithmetic~$\PRA$ proves that for any
formula~$\varphi$ of~$\ZFC$, $\ZFC$ proves~``$(\Set \models \varphi)
\Leftrightarrow \varphi$'', where~``$\Set \models \varphi$'' is to be unrolled
by~$\PRA$.

An alternative is offered by~$\ZFCI$, $\ZFC$ plus the existence of a
strongly inaccessible cardinal. In this system, there is a Grothendieck
universe~$U$; we can form the category~$\Set_U$ of all sets of~$U$ as an honest
set; can define truth in~$\Set_U$; and prove, within the system, that for any
formula~$\varphi$, $(\Set_U \models \varphi) \Leftrightarrow (U \models
\varphi)$. This even holds for formulas of the full infinitary language of
toposes, which allows infinite disjunctions and infinite conjunctions; this
extended language could not be treated by resorting to~$\PRA$ as indicated above.

@@

% XXX: Reference more choice axioms? e.g. Aczel and Rathjen; and that one PhD
% thesis
